---
title: "Assignment 1"
author: "Akos Engelmann, Gergely Paradi, Fabian Gallyas, Ipek Cakin, Simon Jasansky"
date: "2023-03-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


# Load Data
```{r, include=FALSE}
library("tidyverse")
library("ggplot2")
a1_data <- read_csv("a1_data.csv")
print(a1_data)

ggplot(pivot_longer(a1_data, everything()), aes(value)) +
geom_histogram(bins = 100) +
facet_wrap(~name, nrow = 3)


```
```{r}




```


## Define the assumptions that you would like to make.
- There is a causal relationship between the variables
- The causal relationship (i.e. the graph) is acyclic and with no self-loops 

If we use Park algorithm: 
- model is additive noise model
- $\epsilon_1$, $\epsilon_2$ etc. are independent, uniformly distributed with mean 0 and variance $\sigma^2$, and all of the epsilons have to have the same variance. 

However, we cannot assume that the variables themselves are normally distributed! (e.g. look at variable D)


## Develop a procedure to estimate the DAG from the data using your assumptions.




## Implement your procedure in R or Python and report your DAG

## Explain what the limits of your procedure are, both statistically and in terms of possible violations of your assumption with real world data
